---
title: "Anish Ketha - Practice 12 - Fundamental Analysis"
format: html
editor: visual
---

## **Practice 12 Answers** 

#### **Exercise 1**

```{r}
#market timing strat
market_timing_strategy <- function(data, TradingRule, Leverage = 1, Frequency = "Monthly", TradingPeriod){
    # Generate trading signals
    data$RetRf <- data$Ret - data$Rf
    data$TradingRule <- lag(data$TradingRule,1)
    
    data <- na.omit(data[TradingPeriod])

    # Calculate cumulative returns
    data$Strategy_RetRf <- Leverage * data$TradingRule * data$RetRf 
    data$Strategy_Cumulative_Value <- cumprod(1 + data$Strategy_RetRf)  
    
    if (Frequency == "Daily") {
      print(paste('Annualized Excess Return: ', round(mean(data$Strategy_RetRf)*252*100, 2), '%'))
      print(paste('Annualized Volatility: ', round(sd(data$Strategy_RetRf)*sqrt(252)*100, 2), '%'))
      print(paste('Annualized Sharpe Ratio: ', round(mean(data$Strategy_RetRf)/sd(data$Strategy_RetRf)*sqrt(252), 2)))
    } else if (Frequency == "Monthly") {
      print(paste('Annualized Excess Return: ', round(mean(data$Strategy_RetRf)*12*100, 2), '%'))
      print(paste('Annualized Volatility: ', round(sd(data$Strategy_RetRf)*sqrt(12)*100, 2), '%'))
      print(paste('Annualized Sharpe Ratio: ', round(mean(data$Strategy_RetRf)/sd(data$Strategy_RetRf)*sqrt(12), 2)))
    }
    
    plot(index(data), data$Strategy_Cumulative_Value, type = "l", col = "blue", lwd = 2, 
         ylab = "$", xlab = "Date",
         main = paste('Value of $1 invested in the', Frequency, 'Trading Strategy with', Leverage, 'Leverage', TradingPeriod))
    
    return(data)
}

library(quantmod)
spy <- getSymbols("SPY", src = "yahoo", auto.assign = FALSE, from = '1993-01-01', to = '2024-06-30')
spy_ret <- monthlyReturn(Ad(spy))
index(spy_ret) <- as.Date(as.yearmon(index(spy_ret)), frac = 1)
colnames(spy_ret) <- "Ret"

rf <- getSymbols("DFF", src = "FRED", auto.assign = FALSE, from = '1993-01-01', to = '2024-06-30')
# Convert Federal Fund Rates to monthly frequency by dividing 12  
monthly_endpoints <- endpoints(rf, on = "months")
rf <- rf[monthly_endpoints]/12/100
colnames(rf) <- "Rf"

return_data <- merge(spy_ret, rf)
head(return_data, 10)
tail(return_data, 10)

# Load CSV file
econ_indicators_url <- "https://www.dropbox.com/scl/fi/gi2ou5sl1ved66v890p41/Monthly-Market-Return-Predictors.csv?rlkey=15g6j4iel3yxyr810d7cmx7vm&dl=1"

econ_indicators <- read.csv(econ_indicators_url)

# Convert predictors data to xts object
econ_indicators$Date <- as.Date(paste0(econ_indicators$yyyymm, "01"), format = "%Y%m%d")

# Adjust the index of spy_ret to the month end
econ_indicators$Date <- as.Date(as.yearmon(econ_indicators$Date), frac = 1)

# Removes the original yyyymm column
econ_indicators$yyyymm <- NULL

econ_indicators <- xts(econ_indicators[,-ncol(econ_indicators)], order.by = econ_indicators$Date)

head(econ_indicators, 10)

tail(econ_indicators, 10)

#step 2 creating trading signal
default_yield_spread <- econ_indicators$BAA - econ_indicators$AAA
colnames(default_yield_spread) <- "DFY"
default_yield_spread <- na.omit(default_yield_spread)
plot(default_yield_spread, main = "Default Yield Spread", xlab = "Date", ylab = "")

test_data <- merge(return_data, default_yield_spread, join = "inner")
test_data$TradingSignal <- test_data$DFY

head(test_data, 10)
tail(test_data, 10)

#step 3+4 building trading rule and apply backtesting function
test_data$TradingRule <- - test_data$TradingSignal
data_backtest <- market_timing_strategy(data = test_data, TradingRule = TradingRule, Frequency = "Monthly", TradingPeriod = "1993/2024")
```

```{r}
#correct trading rule 
test_data$TradingRule <- test_data$TradingSignal
data_backtest <- market_timing_strategy(data = test_data, TradingRule = TradingRule, Frequency = "Monthly", TradingPeriod = "1993/2024")

#adjust leverage 
data_backtest <- market_timing_strategy(data = test_data, TradingRule = TradingRule, Leverage = 50, Frequency = "Monthly", TradingPeriod = "1993/2024")
```

Why a larger DFY (higher corporate bond risks) predict a higher stock market return in the future?

I think it has to do with business/economic cycles as an increase in DFY may also reflect expectations of higher inflation or rising interest rates. Higher inflation expectations can lead to increased interest rates, which reduce the PV of future cash and once inflation stabilizes, stocks can become more attractive leading to higher stock market returns (and the cycle repeats).
